---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# segregr

<!-- badges: start -->
<!-- badges: end -->

**segregr** calculates spatial and aspatial segregation metrics for areal data
(e.g. census tracts). It implements generalised spatial versions of the
Dissimilarity index, Theil's Entropy and Information Theory index (H),
and the Exposure and Isolation indices. Both global (city-wide scalar) and
local (per areal unit) variants are available. Spatial smoothing is controlled
by a bandwidth parameter using Gaussian or step kernels.

| Index | Global | Local |
|---|---|---|
| Dissimilarity | `D` | `d` |
| Entropy | `E` | — |
| Information Theory (H) | `H` | `h` |
| Isolation | `Q` | `q` |
| Exposure | `P` | `p` |

## Installation

Install the development version from GitHub:

``` r
# install.packages("devtools")
devtools::install_github("mvpsaraiva/segregr")
```

## Usage

### Load data

`measure_segregation()` expects an `sf` object with one row per areal unit,
an `id` column, a `geometry` column, and one column per population group
containing that group's count in each unit.

```{r setup, message = FALSE}
library(sf)
library(segregr)

marilia_sf <- st_read(
  system.file("extdata/marilia_2010.gpkg", package = "segregr"),
  quiet = TRUE
)
head(marilia_sf)
```

### Classical (aspatial) metrics

Pass the sf object to `measure_segregation()`. The default `bandwidths = 0`
returns classical, non-spatial indices.

```{r aspatial}
seg <- measure_segregation(marilia_sf)
```

Global indices are returned as data frames with one row per bandwidth:

```{r global}
seg$D   # Dissimilarity
seg$H   # Information Theory index
seg$E   # Entropy (scalar)
```

The Exposure/Isolation matrix shows how much each group is exposed to every
other group (off-diagonal) and to itself (diagonal = Isolation):

```{r matrix}
exposure_isolation_matrix(seg)
```

Collect all global metrics in one data frame for easy export or comparison:

```{r global-df}
global_metrics_to_df(seg)
```

### Local metrics

Local indices return one value per areal unit. Helper functions join them
back to the input geometries for mapping:

```{r local}
d_sf <- dissimilarity_to_sf(seg)
head(d_sf[, c("id", "bw", "dissimilarity")])
```

`local_metrics_to_sf()` combines all local metrics into a single wide sf
object:

```{r local-all}
local_sf <- local_metrics_to_sf(seg)
names(local_sf)
```

### Spatial metrics with bandwidth

Pass one or more bandwidths (in metres) to smooth population counts across
neighbouring units before computing indices:

```{r spatial}
seg_bw <- measure_segregation(marilia_sf, bandwidths = c(0, 500, 1000, 2000))
global_metrics_to_df(seg_bw)
```

Use `weight_function = "step"` for a binary (moving-window) kernel:

```{r step}
seg_step <- measure_segregation(marilia_sf, bandwidths = 1000,
                                weight_function = "step")
seg_step$D
```

### Mapping local results

```{r map, message = FALSE, warning = FALSE}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  d_bw <- dissimilarity_to_sf(seg_bw, bandwidths = c(0, 1000))

  ggplot2::ggplot(d_bw) +
    ggplot2::geom_sf(ggplot2::aes(fill = dissimilarity), colour = NA) +
    ggplot2::scale_fill_distiller(palette = "Spectral") +
    ggplot2::facet_wrap(~bw, labeller = ggplot2::label_both) +
    ggplot2::theme_void() +
    ggplot2::labs(title = "Local Dissimilarity — Marília (2010)", fill = "d")
}
```

