---
title: "Introduction to segregr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to segregr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

**segregr** calculates spatial and aspatial segregation metrics for areal data
(e.g. census tracts). It provides global (city-wide) and local (per areal unit)
variants of the following indices:

| Index | Global | Local |
|---|---|---|
| Dissimilarity | `D` | `d` |
| Entropy | `E` | (via `h`) |
| Information Theory (H) | `H` | `h` |
| Isolation | `Q` | `q` |
| Exposure | `P` | `p` |

All indices can be computed aspatially (classical form, `bandwidths = 0`) or
with a Gaussian spatial kernel that smooths population counts within a given
bandwidth in metres.

## Installation

```{r, eval = FALSE}
# Install from GitHub
devtools::install_github("mvpsaraiva/segregr")
```

## Loading data

The package ships with a sample dataset from Marília, São Paulo (Brazil),
containing census-tract population counts by race/colour (2010 Census).

```{r setup}
library(sf)
library(segregr)

marilia_sf <- st_read(
  system.file("extdata/marilia_2010.gpkg", package = "segregr"),
  quiet = TRUE
)

head(marilia_sf)
```

The sf object has one row per census tract. Each column after `id` and
`geometry` holds the population count of a group in that tract.

## Aspatial (classical) metrics

Pass the sf object to `measure_segregation()`. The default `bandwidths = 0`
returns the classical, non-spatial indices.

```{r aspatial}
seg <- measure_segregation(marilia_sf)
```

### Global indices

```{r global-d-h}
# Global Dissimilarity index (one row per bandwidth)
seg$D

# Global Entropy
seg$E

# Global Information Theory index H
seg$H
```

The Exposure/Isolation matrix shows how much each group is exposed to every
other group (off-diagonal) and to itself (diagonal = Isolation):

```{r global-matrix}
exposure_isolation_matrix(seg)
```

Collect all global metrics into a single data frame for easy comparison across
bandwidths:

```{r global-df}
global_metrics_to_df(seg)
```

### Local indices

Local indices return one value per areal unit per bandwidth. The helper
functions join these values back to the original geometries for mapping.

```{r local-dissimilarity}
d_sf <- dissimilarity_to_sf(seg)
head(d_sf[, c("id", "bw", "dissimilarity")])
```

```{r local-entropy}
e_sf <- entropy_to_sf(seg)
head(e_sf[, c("id", "bw", "entropy")])
```

```{r local-h}
h_sf <- h_to_sf(seg)
head(h_sf[, c("id", "bw", "h")])
```

```{r local-isolation}
q_sf <- isolation_to_sf(seg)
head(q_sf[, c("id", "bw", "group", "isolation")])
```

```{r local-exposure}
p_sf <- exposure_to_sf(seg)
head(p_sf[, c("id", "bw", "group_a", "group_b", "exposure")])
```

All local metrics can also be joined to the geometries in a single wide sf
object:

```{r local-all}
local_sf <- local_metrics_to_sf(seg)
names(local_sf)
```

## Spatial metrics

Passing one or more bandwidths (in metres) activates Gaussian spatial
smoothing. Population counts are smoothed across neighbouring tracts before
indices are computed, capturing residential mixing at a given spatial scale.

```{r spatial}
seg_spatial <- measure_segregation(marilia_sf, bandwidths = c(0, 500, 1000, 2000))

# Global D at each spatial scale
seg_spatial$D
```

Collect all global metrics at once:

```{r spatial-global-df}
global_metrics_to_df(seg_spatial)
```

### Mapping local indices

```{r maps, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)

d_spatial <- dissimilarity_to_sf(seg_spatial, bandwidths = c(0, 1000))

ggplot(d_spatial) +
  geom_sf(aes(fill = dissimilarity), colour = NA) +
  scale_fill_distiller(palette = "Spectral") +
  facet_wrap(~bw, labeller = label_both) +
  theme_void() +
  labs(title = "Local Dissimilarity index — Marília (2010)",
       fill = "d")
```

```{r maps-h, eval = requireNamespace("ggplot2", quietly = TRUE)}
h_spatial <- h_to_sf(seg_spatial, bandwidths = c(0, 1000))

ggplot(h_spatial) +
  geom_sf(aes(fill = h), colour = NA) +
  scale_fill_distiller(palette = "RdYlGn") +
  facet_wrap(~bw, labeller = label_both) +
  theme_void() +
  labs(title = "Local H index — Marília (2010)", fill = "h")
```

## Step kernel

Use `weight_function = "step"` for a binary kernel that weights all tracts
within the bandwidth equally (equivalent to a moving window average).

```{r step}
seg_step <- measure_segregation(marilia_sf, bandwidths = 1000,
                                weight_function = "step")
seg_step$D
```

## Session info

```{r session}
sessionInfo()
```
